Push-Pull Signal-Function Functional Reactive Programming

### 2012-11-24 Docs/IFL2012/paper.tex
Beginning edits to IFL paper

Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
Recent work on classic FRP has focused on efficient implementation. Reactive is
a system for push-pull FRP~\cite{Elliott2009}. Push-based evaluation evaluates a
system only when input is available, and is thus suitable for discrete inputs
such as events. Pull-based evaluation evaluates the system as quickly as
possible, polling for input, and is preferable for behaviors and signals.
The initial implementations of FRP made use of pull-based evaluation for both
behaviors and events. Reactive, as well as more recent systems such as
``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
events and pull-based evaluation for behaviors.

All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
have used pull-based evaluation. This is due to the ease of implementation of
pull-based evaluation, and the types used for signal functions which do not
permit distinguishing signals and events, or constructing only part of the
input (for instance, one event occurrence.)

@@

\subsection{Examples of Signal Function Primitives}
\label{subsection:Examples_of_Signal_Function_Primitives}
The most basic signal is the {\tt identity} signal function, which, as its
name suggests, simply passes its input to its output.

The {\tt constant} signal function has an empty input, and produces a constant
signal as its output.

The {\tt asap} signal function produces an event at the first time interval
after it begins, and never again.

The {\tt filter} signal function applies a predicate to the values of event
occurrences and passes them along only if the predicate is true of the value.
The predicate outputs {\tt Maybe} values to allow the user to avoid partial
functions on types such as {\tt Maybe} or {\tt Either}.

The {\tt associate} signal function is a routing function, which takes as input
a combination of three signal vectors to which the signal vector append
constructor is combined left-associatively, and produce the same signal vector
as output, but with the signal vector append constructor combined
right-associatively.

The {\tt switch} function is the essential signal function for reactivity. It
acts as the supplied signal function until that signal function produces an
event on its right output. This event's value is a new signal function, which
replaces the switch signal function.

The {\tt loop} function permits a signal function to receive its own output
as an input.

Implementations for these examples will be discussed in Sec.~\ref{subsection:Signal_Function_Implementations}.

@@

The evaluation interfaceis a monad transformer~\cite{Liang1995}. This allows
a signal function evaluation to be constructed from primitive evaluation actions
(event pushing, input signal updating, and sampling) as well as actions from the
underlying monad. These actions can then be run in the underlying monad
to actuate the signal function. Because the actuation takes a signal function's
state as input and produces a new state as output, in addition to the monadic
side effects of handling the signal function's output, different evaluation
actions may be taken on the same signal function from distinct locations within
an application's code. This allows the evaluation interface to be easily
integrated with event-loop style systems as well as traditional imperative
IO systems.

@@

Using GADTs, we can construct representations of signal vectors to use in our
implementation. The first such representation carries a value for each signal
component of a signal vector, and is shown in Figure~\ref{figure:signal_sample_representation}.

\begin{figure}[t]
\begin{code}
data SVSample sv where
  SVSample    :: a -> SVSample (SVSignal a)
  SVSampleEvt :: SVSample (SVEvent a)
  SVNothing   :: SVSample SVEmpty
  SVBoth      :: SVSample svLeft -> SVSample svRight
                 -> SVSample (svLeft :^: svRight)

sample         :: a -> SVSample a
sampleEvt      :: SVSample (SVEvent a)
sampleNothing  :: SVSample SVEmpty
combineSamples :: SVSample svLeft -> SVSample svRight
                  -> SVSample (svLeft :^: svRight)
splitSample    :: SVSample (svLeft :^: svRight)
                  -> (SVSample svLeft, svSample svRight)
sampleValue    :: SVSample (SVSignal a) -> a
\end{code}
\hrule
\caption{Signal sample representation.}
\label{figure:signal_sample_representation}
\end{figure}

A value of this type represents the entire vector, though it has nullary
constructors at event leaves and empty leaves. For event occurrences, we want
to represent only a single point in the vector, as shown in Figure~\ref{figure:event_occurrence_representation}.
This representation uses two constructors, each of which leaves one or the
other of the type variables in the signal vector {\tt SVAppend} constructor
unconstrained.

\begin{figure}[t]
\begin{code}
data SVOccurrence sv where
  SVOccurrence :: a -> SVOccurrence (SVEvent a)
  SVOccLeft    :: SVOccurrence svLeft
                  -> SVOccurrence (svLeft :^: svRight)
  SVOccRight   :: SVOccurrence svRight
                  -> SVOccurrence (svLeft :^: svRight)

occurrence        :: a -> SVOccurrence (SVEvent a)
occLeft           :: SVOccurrence svLeft
                     -> SVOccurrence (svLeft :^: svRight)
occRight          :: SVOccurrence svRight
                     -> SVOccurrence (svLeft :^: svRight)
chooseOccurrences :: SVOccurrence (svLeft :^: svRight)
                     -> Either (SVOccurrence svLeft) (SVOccurrence svRight)
fromOccurrence    :: SVOccurrence (SVEvent a) -> a
\end{code}
\hrule
\caption{Event occurrence representation.}
\label{figure:event_occurrence_representation}
\end{figure}

The final representation ``signal deltas,'' carries replacement values for some
points in a signal vector. Here, we combine the unconstrained type variable
from the occurrence representation with the ``Both'' constructor from the sample
representation. The unconstrained type variable will be placed in a separate
constructor, which allows the construction of empty signal deltas, as is shown
in Figure~\ref{figure:signal_delta_representation}.

\begin{figure}[t]
\begin{code}
data SVDelta sv where
  SVDelta        :: a -> SVDelta (SVSignal a)
  SVDeltaNothing :: SVDelta sv
  SVDeltaBoth    :: SVDelta svLeft -> SVDelta svRight
                    -> SVDelta (svLeft :^: svRight)
delta         :: a -> SVDelta (SVSignal a)
deltaNothing  :: SVDelta sv
combineDeltas :: SVDelta svLeft -> SVDelta svRight
                 -> SVDelta (svLeft :^: svRight)
updateSample  :: SVDelta sv -> SVSample sv -> SVSample sv
\end{code}
\hrule
\caption{Signal delta representation.}
\label{figure:signal_delta_representation}
\end{figure}

>>>>

Classic FRP was first described as an system for interactive animations~\cite{Elliott1997}.
Recent work on classic FRP has focused on efficient implementation. One approach to
efficiency is to separate the evaluation of behaviors and events, since suitable 
strategies give best performance in each case. Push-based evaluation evaluates a
system only when input is available, and is thus suitable for discrete inputs
such as events. Pull-based evaluation evaluates the system as quickly as
possible, polling for input, and is preferable for behaviors and signals.
The initial implementations of FRP made use of pull-based evaluation for both
behaviors and events. Reactive~\cite{Elliott2009}, as well as more recent systems such as
``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
events and pull-based evaluation for behaviors. This is known as ``push-pull''
evaluation.

All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
have used pull-based evaluation for both signals and events. This is due to
the ease of implementation of pull-based evaluation, and the types used for
signal functions which do not permit distinguishing signals and events, or
constructing only part of the input (for instance, one event occurrence.)

@@

@@

The evaluation interface is a monad transformer~\cite{Liang1995}. This allows
a signal function evaluation to be constructed from primitive evaluation actions
(event pushing, input signal updating, and sampling) as well as actions from the
underlying monad. These actions can then be run in the underlying monad
to actuate the signal function. Because the actuation takes a signal function's
state as input and produces a new state as output, in addition to the monadic
side effects of handling the signal function's output, different evaluation
actions may be taken on the same signal function from distinct locations within
an application's code. This allows the evaluation interface to be easily
integrated with event-loop style systems as well as traditional imperative
IO systems.

@@

Using GADTs, we can construct representations of signal vectors to use in our
implementation.

When constructing a representation, we consider how to represent each
component of a signal vector, whether it is a signal element, an empty
element, or an append node. For instance, in order to represent a signal
sample for all signals in an element, we construct the GADT shown in Figure~\ref{figure:signal_sample_representation}.

\begin{figure}
\begin{code}
data SVSample sv where
  SVSample        ::    a -> SVSample (SVSignal a)
  SVSampleEvt     ::    SVSample (SVEvent a)
  SVSampleNothing ::    SVSample SVEmpty
  SVSampleBoth    ::    SVSample svLeft
                     -> SVSample svRight
                     -> SVSample (svLeft :^: svRight)
\end{code}
\hrule
\caption{A representation which carries a value for each signal in a signal vector.}
\label{figure:signal_sample_representation}
\end{figure}
